---
title: "Module 4/5 Project 1"
author: "Annie Dang"
date: "2025-06-12"
output: ioslides_presentation
---

<style type="text/css">
body p, div, h1, h2, h3, h4, h5 {
color: black;
font-family: Modern Computer Roman;
}
slides > slide.title-slide hgroup h1 {
color: #8C1D40; <!--the maroon color-->
}
h2 {
color: #8C1D40; <!-- the maroon color-->
}

slides > slide {
overflow-x: auto !important;
overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Dataset for our Presentation and Analyses

We will use the **Sleep Health and Lifestyle Dataset** on Kaggle for our presentation and analyses.

I picked this dataset because:

- **Personal Relevance**: I tend to stay up and sleep late pretty often and sleep less than seven hours every school day
- **Professional Curiosity**: I am curious about how some parts, such as occupation or the amount of sleep, affect our health
- **Academic Interest**: I am interested in health-related topics whenever I do projects or analyses

## Dataset Quick Summary

Created by Laksika Tharmalingam, the **Sleep Health and Lifestyle Dataset** is a dataset from Kaggle that covers "a wide range of variables related to sleep and daily habits."

- 374 rows, according to the CSV file
- 13 columns: 
 - Person ID, Gender, Age, Occupation
 - Sleep Duration, Quality of Sleep, Physical Activity
 - Stress Level, BMI Category, Blood Pressure
 - Heart Rate, Daily Steps, Sleep Disorder

Link: https://www.kaggle.com/datasets/uom190346a/sleep-health-and-lifestyle-dataset

## Problems

**Occupation Impact by Sleep**: When dealing with stress, it could disrupt our sleep patterns, such as duration and quality. That applies to different occupations, giving some employees work-related stress and irregular schedules. Knowing which occupations are affected can inform workplace wellness policies.

**Question**: "Do certain occupations report significantly shorter sleep duration or low-quality scores?"

**BMI Category and Sleep Health**: While obesity is a known risk factor for sleep disorders, its relationship with perceived sleep quality remains understudied in everyday populations.

**Question**: "How does the BMI category correlate with self-reported sleep quality and stress levels?"

## Load Libraries

Now that we know the problems, we will load some libraries to get the tools required for the dataset analysis. I picked ggplot2, dplyr, and tidyr.

```{r load libraries, echo = TRUE}

# Loading libraries
library(ggplot2)
library(tidyr)
library(dplyr)
```

## Load CSV File

We load the CSV file from the dataset **Sleep_health_and_lifestyle_dataset.csv** and put it in the sleepData variable. That is where we get the rows, columns, and values we need for our analyses.

```{r csv file, echo = TRUE}

# Loading the CSV file
sleepData = read.csv("Sleep_health_and_lifestyle_dataset.csv", 
              header = TRUE)
```

## Check the Variables Summary

To check the loaded CSV file is in place, I use the summary function to view the variables and their information.

```{r variables summary, echo = TRUE}

# Checking that the CSV file is in place
summary(sleepData)
```

## Check the Variables Type

I use the glimpse function to see what variable types the columns are, and we see seven integers, five characters, and one double in those thirteen columns.

```{r variables type, echo = TRUE}

# Checking the variable types
glimpse(sleepData)
```

## Cleaning and Converting Data

We clean the file from the missing values and convert some columns to different variable types.

```{r clean, echo = TRUE}

# Check if the CSV file have missing values
sum(is.na(sleepData))
```

We need to switch the columns from categorical to factorial using the mutate function. That way, it is easier to do analysis testing, such as visualization.

```{r convert, echo = TRUE}

# Converting categorical variables to factors
sleepData = sleepData %>% 
  mutate(
    Occupation = as.factor(Occupation), 
    BMI.Category = as.factor(BMI.Category))
```

## Occupation Impact by Sleep

We want to know if certain occupations report significantly shorter sleep duration or low-quality scores.

```{r occSleep, echo = TRUE}

# Creating the occSleep variable
occSleep = sleepData %>% 
  group_by(Occupation) %>% 
  summarize(
    avgDuration = mean(Sleep.Duration), 
    avgQuality = mean(Quality.of.Sleep)) %>% 
  arrange(avgDuration)
```

## Occupation Impact by Sleep (cont.)

We finished adding some stuff for the occSleep variable, and it is time to create a barplot graph for the chosen columns. We use ggplot to make a graph and put the occSleep variable to extract data we made earlier.

```{r avg duration, echo = TRUE, results = 'hide', fig.show = 'hide'}

# Average Sleep Duration Graph
ggplot(occSleep, aes(x = reorder(Occupation, avgDuration), 
                     y = avgDuration)) + geom_col(fill = "red") + 
  coord_flip() + 
  labs(title = "Average Sleep Duration by Occupation", x = "", 
       y = "Hours") + theme_minimal()
```

We also made a graph for Average Sleep Quality, similar to what we did.

```{r avg quality, echo = TRUE, results = 'hide', fig.show = 'hide'}

# Average Sleep Quality Graph
ggplot(occSleep, aes(x = reorder(Occupation, avgQuality), 
                     y = avgQuality)) + geom_col(fill = "blue") + 
  coord_flip() + 
  labs(title = "Average Sleep Quality by Occupation", x = "", 
       y = "Sleep Quality from 1-10") + theme_minimal()
```

## Occupation Impact by Sleep Graphs

- The first half of these graphs shows surprising similarities, including the y-axis, in contrast to the bottom.
- One of the differences between the graphs is the occupations: nurse, doctor, manager, software engineer, and teacher switch different positions.
- Engineers have longer sleep duration and better sleep quality than sales representatives

```{r avg sleep duration and quality, fig.show = 'asis'}

# Average Sleep Duration Graph
ggplot(occSleep, aes(x = reorder(Occupation, avgDuration), 
                     y = avgDuration)) + geom_col(fill = "red") + 
  coord_flip() + 
  labs(title = "Average Sleep Duration by Occupation", x = "", 
       y = "Hours") + theme_minimal()

# Average Sleep Quality Graph
ggplot(occSleep, aes(x = reorder(Occupation, avgQuality), 
                     y = avgQuality)) + geom_col(fill = "blue") + 
  coord_flip() + 
  labs(title = "Average Sleep Quality by Occupation", x = "", 
       y = "Sleep Quality from 1-10") + theme_minimal()
```

## BMI Category and Sleep Health

We want to know how the BMI category correlates with self-reported sleep quality and stress levels. That is like the first problem, except we do not create a custom variable. Also, instead of a barplot, we use a boxplot graph and add a legend.

```{r bmi category and sleep health, echo = TRUE, results = 'hide', fig.show = 'hide'}

# Sleep Quality Graph
ggplot(sleepData, aes(x = BMI.Category, y = Quality.of.Sleep)) + 
  geom_boxplot(aes(fill = BMI.Category)) + 
  labs(title = "Sleep Quality by BMI Category", x = "BMI Category", 
       y = "Sleep Quality (1-10)", fill = "BMI Category") + 
  theme_minimal()

# Stress Level Graph
ggplot(sleepData, aes(x = BMI.Category, y = Stress.Level)) + 
  geom_boxplot(aes(fill = BMI.Category)) + 
  labs(title = "Stress Level by BMI Category", x = "BMI Category", 
       y = "Stress Level (1-10)", fill = "BMI Category") + 
  theme_minimal()
```

## BMI Category and Sleep Health Graphs

- The x-axis shows different categories: normal, normal weight, obese, and overweight. Each has varied colors, which you can see on the legend.
- The normal BMI categories have better sleep quality and lower stress level than the overweight and obese categories in terms of medians and interquartile range

```{r sleep quality and stress level, fig.show = 'asis'}

# Sleep Quality Graph
ggplot(sleepData, aes(x = BMI.Category, y = Quality.of.Sleep)) + 
  geom_boxplot(aes(fill = BMI.Category)) + 
  labs(title = "Sleep Quality by BMI Category", x = "BMI Category", 
       y = "Sleep Quality (1-10)", fill = "BMI Category") + 
  theme_minimal()

# Stress Level Graph
ggplot(sleepData, aes(x = BMI.Category, y = Stress.Level)) + 
  geom_boxplot(aes(fill = BMI.Category)) + 
  labs(title = "Stress Level by BMI Category", x = "BMI Category", 
       y = "Stress Level (1-10)", fill = "BMI Category") + 
  theme_minimal()
```

## Conclusion

### Occupation Impact on Sleep
- Engineers have the highest sleep duration (8 hours) and best sleep quality (8.5/10)
- Sales Representatives have the lowest sleep duration (around 5.8 hours) and poorest sleep quality (4/10), suggesting high work-related stress
- Actionable insight: High-stress occupations, such as sales or scientists, may benefit from workplace wellness programs to improve sleep

### BMI Category and Sleep Health
- Employees with normal and normal weight BMI report better sleep quality (Median: 8/10) and lower stress (Median: 5-6/10)
- Employees with overweight and obese show poorer sleep quality (Median: 6-7/10) and higher stress (Median: 6-7)
- Actionable insight: Weight management initiatives could enhance both sleep and mental health

## Limitations and Future Work

### Limitations
- Possible biases in self-reported data
- Uneven sample size per occupation and BMI group

### Future Work
- Investigate specific interventions
- Examples: Flexible schedules for nurses, stress workshops for the sales team, etc.